---
title: "Homework 02 - Build better data"
subtitle: "INFO 4940/5940 - Fall 2025"
author: "Paul Vermette II"
date: today
format: typst
engine: knitr
---

# Setup

Load packages and data:

```{r}
library(tidyverse)
library(tidymodels)

set.seed(123)

scorecard <- read_csv("data/scorecard.csv")

glimpse(scorecard)
```



# Exercise 1
```{r}
# checking for missing values in outcome variable 'debt'

missing_debt <- sum(is.na(scorecard$debt))
cat("Missing values in 'debt' variable:", missing_debt, "\n")
```

```{r}
# Removing rows with missing outcome variable :)

scorecard_clean <- scorecard |>
  filter(!is.na(debt))

cat("Rows after removing missing debt:", nrow(scorecard_clean), "\n")
cat("Rows removed", nrow(scorecard) - nrow(scorecard_clean), "\n")

summary(scorecard_clean$debt)

# I did this because dropping missing outcomes BEFORE splitting avoids leakage. It also prevents the resampling from being shaped by rows that wont be usable down the road
```



```{r}
# particioning!

set.seed(123)

data_split <- initial_split(
  scorecard_clean,
  prop = 0.75,
  strata = debt
)

# I chose a 75/25 split because it's commonly practiced in TMWR examples. We have plenty of data to train, and enough holdout for a solid final estimate
```

```{r}
# extracting training and testing sets

scorecard_train <- training(data_split)
scorecard_test <- testing(data_split)
```

```{r}
# Verifying the results

cat("\n=== Data Split Summary ===\n")
cat("Training set:", nrow(scorecard_train), "observations\n")
cat("Test set:", nrow(scorecard_test), "observations\n")
cat(
  "Training proportion:",
  round(nrow(scorecard_train) / nrow(scorecard_clean), 3),
  "\n"
)
```



```{r}
# Verifying stratification worked

train_summary <- summary(scorecard_train$debt)
test_summary <- summary(scorecard_test$debt)

cat("\n=== Outcome Distribution Comparison ===\n")
cat("Training set debt summary:\n")
print(train_summary)
cat("\nTest set debt summary:\n")
print(test_summary)
```


```{r}
cat("\nKey Statistics Comparison:\n")
cat(
  "Training - Mean:",
  round(mean(scorecard_train$debt), 0),
  ", Median:",
  round(median(scorecard_train$debt), 0),
  "\n"
)
cat(
  "Test - Mean:",
  round(mean(scorecard_test$debt), 0),
  ", Median:",
  round(median(scorecard_test$debt), 0),
  "\n"
)

```


```{r}
# Visualization

debt_comparison <- bind_rows(
  scorecard_train |> select(debt) |> mutate(split = "Training"),
  scorecard_test |> select(debt) |> mutate(split = "Testing")
)

ggplot(debt_comparison, aes(x = debt)) +
  geom_histogram(bins = 30, alpha = 0.8, color = "white", fill = "blue") +
  facet_wrap(~split, ncol = 1) +
  labs(
    title = "Distribution of Student Debt by Data Split",
    subtitle = "Verifying stratification maintained similar distributions",
    x = "Median Student Debt ($)",
    y = "Number of Institutions"
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal()

# The training and testing distributions are visually aligned. For our model, it confirms that our split was properly stratified!
```

# Justification! 

```{r}
# The 75/25 split is shown in many examples from our readings, as it provides substantial training while preserving adqeuate test date for reliable performance estimation. In ISLR ch5, they emphasize the importance of sufficient test data for validation.

# Stratified sampling is recommeneded on the outcome variable because then the training and test sets have similar target distributions. It will prevent a biased performance that oculd occur if one set contained disproportionately high or low debt institutions

# For the missing value handling before we split, they mention in TMRW ch4 that decisions invilving an outcome variable should be made before data pslitting to avoid data leakage.

# set.seed(123) makes sure we have a consistent split across analysis

# For the sole purpose of this assignment, we are just using a simple train/test split rather than train/validation/test. Cross-validation will be used in subsequent exercises (I think) for model tuning and selection
```

# Exercise 2

```{r}
library(tidyverse)
library(corrplot)
library(GGally)
```

```{r}
# Using ONLY training data for Exploratory Analysis

exploratory_data <- scorecard_train # to avoid using data from test set to guide any further decisions
```


```{r}
# Basic Summary Stats

cat("=== DEBT DISTRIBUTION ANALYSIS ===\n")
debt_summary <- summary(exploratory_data$debt)
print(debt_summary)

cat("\nAdditional Statistics:\n")
cat("Standard Deviation:", round(sd(exploratory_data$debt), 2), "\n")
cat("IQR:", round(IQR(exploratory_data$debt), 2), "\n")
cat(
  "Coefficient of Variation:",
  round(sd(exploratory_data$debt) / mean(exploratory_data$debt), 3),
  "\n"
)

#analysis: the mean and median differ by just $77.
## Moderate spread (CV ~ .32) will give our model sufficient variability to learn from.
# High debt outliers will need residual checks
```


```{r}
# Distribution Viz

p1 <- ggplot(exploratory_data, aes(x = debt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(
    aes(xintercept = mean(debt)),
    color = "red",
    linetype = "dashed",
    size = 1
  ) +
  geom_vline(
    aes(xintercept = median(debt)),
    color = "orange",
    linetype = "dashed",
    size = 1
  ) +
  labs(
    title = "Distribution of Student Debt (Training Data)",
    subtitle = "Red = Mean, Orange = Median",
    x = "Median Student Debt ($)",
    y = "Count"
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal()

```


```{r}
# Box plot

p2 <- ggplot(exploratory_data, aes(y = debt)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  labs(
    title = "Box Plot of Student Debt",
    y = "Median Student Debt ($)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()
```

```{r}
# QQ plot!

p3 <- ggplot(exploratory_data, aes(sample = debt)) +
  geom_qq() +
  geom_qq_line(color = "red") +
  labs(
    title = "Q-Q Plot: Debt Distribution vs Normal",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal()
```

```{r}
# Actual Visualizations

print(p1)
print(p2)
print(p3)
```

# Analysis 

```{r}
## Image 1 - Histogram:
# The roughly normal distribution (slight right skew) is excellent for linear regression, since it meets the normality assumption without requiring transformation. Our models should predict well for typical institutions, but the extended right tail indicates some high debt cases that may be harder to predict accurately.

## Image 2 - Box Plot
# The clear outliers above $30,000 will likely influence our model fit and may require hardcore regression methods or outlier investigation during model diagnostics. Feature engineering around debt ranges/categories could be valuable

## Image 3 - Q-Q Plot
# EXCELLENT normality in the middle quartiles! That confirms that OLS regression assumptions are met for most institutions, which means we can trust standard linear model inference and diagnostics. The upper tail deviations indicate that we should monitor residual plots for those high-debt institutions in model eval.

## Overall:
# There are patterns in distribution that support starting with linear regression as our baseline. Although, we should prepare to handle outliers through more complex methods and consider the slight skew when we interpret model/prediction intervals.

```

# Exercise 3

```{r}
# Training Data Assessment

set.seed(123)

nrow(scorecard_train)

cv_folds <- vfold_cv(
  scorecard_train,
  v = 10, #standard in ISL/R for bias-variance tradeoff
  strata = debt
)

#Justification: a 10-fold balances bias/variance and is the best choice via our reading in ISLR chapter 5.
```

```{r}
# Verifying it worked

fold_summary <- map_dfr(cv_folds$splits, function(split) {
  tibble(
    train_mean = mean(analysis(split)$debt),
    test_mean = mean(assessment(split)$debt)
  )
})

summary(fold_summary)

#analysis: Training and assessment folds have nearly identical means with a difference of ~$1
# tight ranges show that stratification on debt works perfectly. so, CV estimates are unbiased
```


```{r}
# Null model

null_spec <- null_model() |>
  set_engine("parsnip") |>
  set_mode("regression")

null_workflow <- workflow() |>
  add_model(null_spec) |>
  add_variables(outcomes = debt, predictors = c())

set.seed(123)
null_cv_results <- fit_resamples(
  null_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq),
  control = control_resamples(save_pred = TRUE)
)

# Justification: Tidymodels' null model predicts the training mean, so there's no formula used. This is the right baseline to beat. The 'rsq' is NA on purpose because it helps make predictions more constant.
```

```{r}
# Collect and show metrics

null_metrics <- collect_metrics(null_cv_results)
null_metrics


```

```{r}
# baseline comparison of mean

null_rmse <- null_metrics |>
  filter(.metric == "rmse") |>
  pull(mean)

null_mae <- null_metrics |>
  filter(.metric == "mae") |>
  pull(mean)

null_rsq <- null_metrics |>
  filter(.metric == "rsq") |>
  pull(mean)

```

```{r}
# What does the null model predict?

training_mean <- mean(scorecard_train$debt)
training_mean
```



```{r}
# Baseline metrics

null_rmse

null_mae

sd(scorecard_train$debt)

null_rmse / sd(scorecard_train$debt)

#analysis: RMSE/SD ratio = 0.999, which confirms that this is a true mean-only baseline!
```


# Analysis 
```{r}
# The null model shows a clear baseline for comparison. It has an RMSE of $4,873 and a MAE of $3,884, by simply predicting the training mean ($15,077) for all institutions. The RMSE to standard deviation ratio of 0.999 confrims that this is functioning as a legitimate null model. The undefined R^2 (NaN) is expeted since all predictions are identical.

# The baseline represents the 'worst acceptable performance'; any meaningful predictive model must substantially beat these metrics! Given that the average prediction error is $4,000, there's a lot of room for improvement, and our exploratory analysis suggests strong relationships between debt and institutional characteristics, like costs and earnings. This indicates that linear models should achieve meaninfdul reductions in prediction error.
```


# Exercise 4

```{r}
library(tidymodels)
```

```{r}
#establish

set.seed(123)

lm_spec <- linear_reg() |>
  set_engine("lm") |>
  set_mode("regression")
```

```{r}
# preprocessing recipe

basic_recipe <- recipe(debt ~ ., data = scorecard_train) |>
  step_rm(unit_id, name) |>
  step_impute_mean(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_corr(all_numeric_predictors(), threshold = 0.9)
```

```{r}
# workflow

basic_lm_workflow <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(basic_recipe)
```

```{r}
# fit for cross-val

set.seed(123)
basic_lm_cv_results <- fit_resamples(
  basic_lm_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq),
  control = control_resamples(save_pred = TRUE)
)

```

```{r}
# metrics

basic_lm_metrics <- collect_metrics(basic_lm_cv_results)
basic_lm_metrics

# Analysis: This is an improved outcome vs. null
```

```{r}
lm_rmse <- basic_lm_metrics |>
  filter(.metric == "rmse") |>
  pull(mean)

lm_mae <- basic_lm_metrics |>
  filter(.metric == "mae") |>
  pull(mean)

lm_rsq <- basic_lm_metrics |>
  filter(.metric == "rsq") |>
  pull(mean)
```

```{r}
# compare to null model

lm_rmse
lm_mae
lm_rsq
```

```{r}
# improvements calculated

rmse_improvement <- (null_rmse - lm_rmse) / null_rmse * 100
mae_improvement <- (null_mae - lm_mae) / null_mae * 100

rmse_improvement
mae_improvement
```

```{r}
# null model baseline (reference)

null_rmse
null_mae
```


# Exercise 5

```{r}
# Feature Engineering

set.seed(123)

lm_spec <- linear_reg() |>
  set_engine("lm") |>
  set_mode("regression")

``` 



```{r}
# base recipe

base_recipe <- recipe(debt ~ ., data = scorecard_train) |>
  step_rm(unit_id, name) |>
  step_impute_mean(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_corr(all_numeric_predictors(), threshold = 0.9)

```


# I am chosing to do one test at a time first. I don't want to throw a ton of features in at once. I would rather prevent overfitting. So, I created 5 targeted candidates derived from earlier EDA and tested them individually with CV! 

```{r}
# test 1: cost to earnings ratio
# rationale: hypothesized driver of debt burden (hurt performance, dropped)
test1_recipe <- base_recipe |>
  step_mutate(cost_earnings_ratio = cost_net / earnings_med) |>
  step_mutate(cost_earnings_ratio = pmin(cost_earnings_ratio, 5, na.rm = TRUE))

test1_workflow <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(test1_recipe)

set.seed(123)
test1_results <- fit_resamples(
  test1_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq)
)

test1_metrics <- collect_metrics(test1_results)
test1_rmse <- test1_metrics |> filter(.metric == "rmse") |> pull(mean)
test1_mae <- test1_metrics |> filter(.metric == "mae") |> pull(mean)

```

```{r}
# test1 results:
test1_rmse
test1_mae
cat(
  "Test 1 improvement - RMSE:",
  round((lm_rmse - test1_rmse) / lm_rmse * 100, 2),
  "%\n"
)

#analyze: No improvement..
```

```{r}
# test 2: affordability ratio (household income vs cost: plausible affordability signal)

test2_recipe <- base_recipe |>
  step_mutate(affordability_ratio = median_hh_inc / cost_net) |>
  step_mutate(affordability_ratio = pmin(affordability_ratio, 10, na.rm = TRUE))

test2_workflow <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(test2_recipe)

set.seed(123)
test2_results <- fit_resamples(
  test2_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq)
)

test2_metrics <- collect_metrics(test2_results)
test2_rmse <- test2_metrics |> filter(.metric == "rmse") |> pull(mean)
test2_mae <- test2_metrics |> filter(.metric == "mae") |> pull(mean)

# Test 2 results
test2_rmse
test2_mae
cat(
  "Test 2 improvement - RMSE:",
  round((lm_rmse - test2_rmse) / lm_rmse * 100, 2),
  "%\n"
)

#analyze: STRONGLY degrades performance, it's likely unstable because of divisions by small cost values and high variance in household income
```


```{r}
# test 3: comp_rate * retention_rate, which is an interaction reflecting institutional performance.

test3_recipe <- base_recipe |>
  step_mutate(academic_success = comp_rate * retention_rate)

test3_workflow <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(test3_recipe)

set.seed(123)
test3_results <- fit_resamples(
  test3_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq)
)

test3_metrics <- collect_metrics(test3_results)
test3_rmse <- test3_metrics |> filter(.metric == "rmse") |> pull(mean)
test3_mae <- test3_metrics |> filter(.metric == "mae") |> pull(mean)

# Test 3 results
test3_rmse
test3_mae
cat(
  "Test 3 improvement - RMSE:",
  round((lm_rmse - test3_rmse) / lm_rmse * 100, 2),
  "%\n"
)

#analyze: Tiny, somewhat negligible, positive effect. There's some school-quality signal but not anywhere near enough to matter
```

```{r}
# Test 4: Economic need indicator (Pell + first generation)
# Justification: socioeconomic load on borrowing

test4_recipe <- base_recipe |>
  step_mutate(economic_need = pell_pct + first_gen)

test4_workflow <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(test4_recipe)

set.seed(123)
test4_results <- fit_resamples(
  test4_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq)
)

test4_metrics <- collect_metrics(test4_results)
test4_rmse <- test4_metrics |> filter(.metric == "rmse") |> pull(mean)
test4_mae <- test4_metrics |> filter(.metric == "mae") |> pull(mean)

# Test 4 results
test4_rmse
test4_mae
cat(
  "Test 4 improvement - RMSE:",
  round((lm_rmse - test4_rmse) / lm_rmse * 100, 2),
  "%\n"
)

# Analyze: No measureable effect
```



```{r}
# test 5: nonlinearity in costs' effect on debt (helped - kept)
test5_recipe <- base_recipe |>
  step_mutate(cost_net_sq = cost_net^2) |>
  step_normalize(cost_net_sq)

test5_workflow <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(test5_recipe)

set.seed(123)
test5_results <- fit_resamples(
  test5_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq)
)

test5_metrics <- collect_metrics(test5_results)
test5_rmse <- test5_metrics |> filter(.metric == "rmse") |> pull(mean)
test5_mae <- test5_metrics |> filter(.metric == "mae") |> pull(mean)

# Test 5 results
test5_rmse
test5_mae
cat(
  "Test 5 improvement - RMSE:",
  round((lm_rmse - test5_rmse) / lm_rmse * 100, 2),
  "%\n"
)

# analyze: ONLY feature with a consistent gain. A quadratic term captures the mild non linearity in how very high costs can drive debt!
```

```{r}
final_enhanced_recipe <- base_recipe |>
  step_mutate(
    cost_net_sq = cost_net^2,
    academic_success = comp_rate * retention_rate
  ) |>
  step_normalize(cost_net_sq)

# justification
# i only combined the features that helped or were neutral. The result was small, but a very real improvement.
```


```{r}
final_enhanced_workflow <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(final_enhanced_recipe)
```


```{r}
# Fit final enhanced model using cross-validation
set.seed(123)
final_enhanced_results <- fit_resamples(
  final_enhanced_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq),
  control = control_resamples(save_pred = TRUE)
)

```

```{r}
# Collect final results
final_enhanced_metrics <- collect_metrics(final_enhanced_results)
final_enhanced_metrics

# Analyze: the final result showed a small but real improvement over basic LM The gain mainly came from the quadratic cost term (cost_net_sq). All other features engineered had no effect or even worsened performance.
```

```{r}
final_rmse <- final_enhanced_metrics |> filter(.metric == "rmse") |> pull(mean)
final_mae <- final_enhanced_metrics |> filter(.metric == "mae") |> pull(mean)
final_rsq <- final_enhanced_metrics |> filter(.metric == "rsq") |> pull(mean)


final_rmse
final_mae
final_rsq

#analyze: The change is solid here! There a .58% decrease of RMSE, and a 00.98% decrease in MAE, plus 0.6ppt of R^2

```


```{r}
final_rmse_improvement <- (lm_rmse - final_rmse) / lm_rmse * 100
final_mae_improvement <- (lm_mae - final_mae) / lm_mae * 100
```



```{r}
# Feature Engineering Justification!

# 1. Cost ratios capture debt burden relative to benefits
# 2. Academic metrics indicate insitutional quality or value
# 3. Socioeconomic factors reflect ability to pay
# 4. Geographic variables account for regional cost differences
# 5. Ratio capping prevents EXTREME outliers from dominating the entire model
```





# Exercise 6

```{r}
# Random Forest Baseline

# The model is a random forest with the ranger engine
# Resampling a 10-fold CV on the same stratified folds
# trees = 1000: large forest for variance reduction! It is supposed to be cheap with ranger
# the 'mtry/min_n' tuned lets the CV pick feature resampling + terminal node sizes for best bias-variance tradeoff

set.seed(123)

rf_spec <- rand_forest(
  trees = 1000,
  mtry = tune(),
  min_n = tune()
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")
```



```{r}
# Processing trees
# step_rm: Id/label columns carrying no generalizable signal.
# no dummies/scaling because ranger handles factors and scale-free splits

rf_recipe <- recipe(debt ~ ., data = scorecard_train) |>
  step_rm(unit_id, name) |>
  step_impute_mean(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors())

rf_wf <- workflow() |>
  add_model(rf_spec) |>
  add_recipe(rf_recipe)

```

```{r}
# Tuning grid

# finalizing predictors.
# min_n 2...20 controls the node size

rf_grid <- grid_regular(
  finalize(mtry(), scorecard_train |> select(-debt)),
  min_n(range = c(2L, 20L)),
  levels = 5
)

set.seed(123)
rf_tuned <- tune_grid(
  rf_wf,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(rmse, mae, rsq),
  control = control_grid(save_pred = TRUE)
)

rf_metrics <- collect_metrics(rf_tuned)
rf_metrics

```

```{r}
glimpse(rf_metrics)

```




# Exercise 7

```{r}
best_rf <- select_best(rf_tuned, metric = "rmse")
rf_final_wf <- finalize_workflow(rf_wf, best_rf)

# select_best picks the hyper parameters with lowest mean CV RMSE.
# Finalize_workflow locks spec+recipe with those values
```

```{r}
set.seed(123)
rf_test_fit <- last_fit(
  rf_final_wf,
  split = data_split,
  metrics = metric_set(rmse, mae, rsq)
)
collect_metrics(rf_test_fit)

# Analysis: CLEARLY outperforms every linear model! It will be the final selected model. It exceeds the 25% goal!
```

# Analysis: The best fit was a tuned random forest. We had test set metrics of: RMSE = 3,218, MAE = 2,303. We improved vs the null model (RMSE = 4,873, MAE = 3,884) and basic linear model (RMSE = 3,652, MAE = 2,733). The linear regression models didn't do badly either. Basic LM (RMSE = 3,652, MAE = 2,733). Enhanced LM (RMSE = 3,631, MAE = 2,706). So, feature engineering here only had ~1% improvement over the basic LM. This confirms that the linear model was already near its limit. 

# further analysis: The dataset contains nonlinear relationships and feature interactions that linear models can't capture well. Random forest natraully models these complexities, which offers a CLEAR and substantial performance edge. 

# Final statement: the tuned random forest model is the best chosen predictor of median student debt. It exceeds the 25% reduction goal relative to the null model baseline and significantly out performs every linear alternative. This is the best performing model for this homework. 

# GAI self-reflection 
# I had to use Claude to help me connect some of the analysis at the end. I specifically struggled in how to interpret some of the test results, but after Claude explained it to me as I asked it to 'be my professor, not homework helper', it guided me instead of giving me the answer. 


