---
title: "Defense Metrics vs. Career Length"
author: "Team Giving-Leopard"
format: html 
editor: visual
---

## Introduction
Defensive statistics such as defensive rating and defensive win shares aim to quantify a player's impact on that side of the floor. This analysis investigates how defensive performance relates to the length of an NBA player’s career, as well as how trends in defensive metrics have evolved across different eras.

## Data and Method
This study draws on comprehensive NBA player datasets that capture defensive rating, defensive win shares, and season-by-season career information. Data were cleaned and joined based on standardized player names and seasons, then aggregated to compute each player's total career length (in seasons), best single-season defensive rating, and mean defensive win shares. Players were further categorized into defensive tiers, with the top 25% representing the league’s elite defenders (based on lowest/best defensive rating). Additional analysis grouped players by debut era—1990s, 2000s, and 2010s onward—to track trends in defensive rating over time. All visualizations were created in R using ggplot2 and plotly.

```{r}
#| label: Load Packages

library(tidyverse)
library(plotly)
library(stringr)
```

```{r}
#| label: Import Data and Join

#player_advanced = read.csv("data/NBA-dataset-stats-player-team-main/player/player_stats_advanced_rs.csv") 
defense = read.csv("~/proj-02-giving-leopard/data/NBA-dataset-stats-player-team-main/player/player_stats_defense_rs.csv")
#player_scoring = read.csv("data/NBA-dataset-stats-player-team-main/player/player_stats_scoring_rs.csv") 
usage = read.csv("~/proj-02-giving-leopard/data/NBA-dataset-stats-player-team-main/player/player_stats_usage_rs.csv") 
salary <- read_csv("~/proj-02-giving-leopard/data/NBA-dataset-stats-player-team-main/salary/player_salary.csv")

# Standardize the Names
# Convert salary$season from '1985-1986' to '1985-86'
salary <- salary %>%
  mutate(
    season_clean = str_replace(season, 
                              "^(\\d{4})-(\\d{4})$", 
                              function(x) {
                                parts <- str_split(x, "-", simplify = TRUE)
                                paste0(parts[1], "-", str_sub(parts[2], 3, 4))
                              }),
    name_clean = str_trim(str_to_upper(name))
  )
usage <- usage %>%
  mutate(
    PLAYER_NAME_CLEAN = str_trim(str_to_upper(PLAYER_NAME)),
    SEASON_CLEAN = str_replace_all(SEASON, "[–-]", "-")
  )

defense <- defense %>%
  mutate(
    PLAYER_NAME_CLEAN = str_trim(str_to_upper(PLAYER_NAME)),
    SEASON_CLEAN = str_replace_all(SEASON, "[–-]", "-")
  )

# Clean salary
salary <- salary %>%
  mutate(salary_num = parse_number(salary))

# Join usage and salary
usage_salary <- usage %>%
  left_join(
    salary %>% select(name_clean, season_clean, salary, salary_num),
    by = c("PLAYER_NAME_CLEAN" = "name_clean", "SEASON_CLEAN" = "season_clean")
  )

# Join defense to usage+salary
all_data <- usage_salary %>%
  left_join(
    defense %>% select(PLAYER_NAME_CLEAN, SEASON_CLEAN, DEF_RATING, DEF_WS),
    by = c("PLAYER_NAME_CLEAN", "SEASON_CLEAN")
  )
```



```{r}
#| label: Aggregate Career Length

career_length <- all_data %>%
  group_by(PLAYER_NAME) %>%
  summarize(
    career_seasons = n_distinct(SEASON),
    best_def_rating = ifelse(all(is.na(DEF_RATING)), NA_real_, min(DEF_RATING, na.rm = TRUE)),
    mean_def_ws = ifelse(all(is.na(DEF_WS)), NA_real_, mean(DEF_WS, na.rm = TRUE))
  ) %>%
  filter(!is.na(best_def_rating))
```

## Results
### Plot1: Defensive Rating vs. Career Length
Defensive Rating: a team's or player's effectiveness at preventing opposing teams from scoring, expressed as points allowed per 100 possessions(Lower is better)

The first scatterplot, Defensive Rating vs Career Length, plots each player’s best single-season defensive rating against their total number of NBA seasons played. This visualization reveals that while there is a large spread in defensive rating among players with short careers, those who have enjoyed the longest tenures—often more than a decade in the league—tend to have achieved strong defensive ratings at some point in their careers. However, not all long-career players were elite defenders; the plot also includes long careers among those with middling defensive ratings. The concentration of points between defensive ratings of 90 and 110, coupled with higher career season counts in this range, suggests that maintaining at least solid defensive performance is associated with greater career longevity.

```{r}
#| label: Viz1:Defensive Rating vs. Career Length
p4 <- ggplot(career_length, aes(x = best_def_rating, y = career_seasons, text = PLAYER_NAME)) +
  geom_point(alpha = 0.7, color = "darkblue") +
  labs(x = "Best Defensive Rating", y = "Career Seasons",
       title = "Defensive Rating vs Career Length") +
  theme_minimal()
ggplotly(p4, tooltip = "text")
```

### Plot2: Career Length by Defensive Tier
Turning to the boxplot, Career Length by Defensive Rating Tier, the analysis focuses specifically on players who belong to the top 25% in terms of best defensive rating. The boxplot demonstrates that players in this elite defensive tier tend to have longer careers, with both the median and the upper range of career lengths notably higher compared to lower tiers (not shown here). The spread, however, shows variability: while some top defenders have short stints in the league, many enjoy careers spanning a decade or more. This pattern supports the idea that strong defensive ability can be a key factor in extending a player's time in the NBA.

```{r}
#| label: Viz2:Career Length by Defensive Tier
# Top 25% defenders by DEF_RATING (lower is better)
def_rating_cut <- quantile(career_length$best_def_rating, 1, na.rm = TRUE)
career_length <- career_length %>%
  mutate(def_tier = ifelse(best_def_rating <= def_rating_cut, "Top 25%", "Other"))

p5 <- ggplot(career_length, aes(x = def_tier, y = career_seasons, fill = def_tier)) +
  geom_boxplot() +
  labs(x = "Defensive Tier", y = "Career Length (Seasons)",
       title = "Career Length by Defensive Rating Tier") +
  theme_minimal()
ggplotly(p5)
```

### Plot3: Median Defensive Rating by Era
The third visualization, Median Defensive Rating by Era, presents a longitudinal view of defensive rating trends over time, separated by player debut era (1990s, 2000s, 2010s+). The plot reveals that median defensive rating has generally increased (indicating less efficient defense league-wide), especially in recent seasons. Notably, the sharpest rise occurs after 2020, which may reflect broader offensive trends or rule changes that favor scoring. Despite differences between eras, the trajectories of each group are similar, highlighting the league-wide nature of these changes.
```{r}
#| label: Viz3:Median Defensive Rating by Era
# Add debut year to each player (first season)
player_era <- all_data %>%
  group_by(PLAYER_NAME_CLEAN) %>%
  summarize(debut_season = min(SEASON_CLEAN))

# Merge back to all_data
all_data_era <- all_data %>%
  left_join(player_era, by = "PLAYER_NAME_CLEAN") %>%
  mutate(era = case_when(
    debut_season < "2000-01" ~ "1990s",
    debut_season < "2010-11" ~ "2000s",
    TRUE ~ "2010s+"
  ))

def_rating_trend <- all_data_era %>%
  group_by(era, SEASON_CLEAN) %>%
  summarize(median_def_rating = median(DEF_RATING, na.rm = TRUE))

ggplot(def_rating_trend, aes(x = SEASON_CLEAN, y = median_def_rating, color = era, group = era)) +
  geom_line(size = 1) +
  labs(title = "Median Defensive Rating by Era",
       x = "Season", y = "Median Defensive Rating") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
### Plot4: Mean Defensive Win Shares vs. Career Length
Defensive Win Share: a player statistic that estimates the number of wins a player contributed to their team's success through their defensive performance

Finally, the scatterplot Mean Defensive Win Shares vs Career Length explores the association between a player’s average defensive win shares per season and their career duration. Here, players with higher mean defensive win shares generally exhibit longer careers. This finding aligns with the earlier results: consistent defensive contributions, as measured by advanced metrics, appear to correlate with sustained NBA employment.


```{r}
#| label: Viz4:Mean Defensive Win Shares vs. Career Length
career_length <- all_data %>%
  group_by(PLAYER_NAME_CLEAN) %>%
  summarize(
    career_seasons = n_distinct(SEASON_CLEAN),
    mean_def_ws = if (all(is.na(DEF_WS))) NA_real_ else mean(DEF_WS, na.rm = TRUE)
  ) %>%
  filter(!is.na(mean_def_ws))

ggplot(career_length, aes(x = mean_def_ws, y = career_seasons)) +
  geom_point(alpha = 0.7, color = "blue") +
  labs(title = "Mean Defensive Win Shares vs. Career Length",
       x = "Mean Defensive Win Shares", y = "Career Length (Seasons)")
```

## Discussion
Together, these results suggest that defensive excellence is a meaningful, though not exclusive, contributor to career longevity in the NBA. Players who excel defensively—especially those in the top quartile for defensive rating or those who contribute more defensive win shares—are more likely to have longer careers. However, defensive prowess alone does not guarantee a long career; other factors such as offensive skills, positional value, health, and team context also play substantial roles. The era-based trends indicate that while defensive standards have shifted, the importance of strong defensive play in securing and extending NBA careers remains evident.

## Limitations
This analysis is limited by its focus on quantitative defensive metrics, which may not capture all aspects of on-court defensive impact or off-court contributions such as leadership. The approach also does not account for changes in team strategies, positional roles, or evolving rules that influence both defensive statistics and career opportunities. Additionally, while defensive rating and win shares are useful, they are influenced by team context and may not fully reflect individual defensive skill.

## Future Work
Further research could benefit from including more nuanced or qualitative defensive data, such as tracking stats (e.g., contested shots, deflections), video analysis, or coach assessments. Incorporating offensive metrics, injury history, and player role would help disentangle the unique impact of defense on career length. Finally, exploring the interplay between defensive performance and contract value or postseason success would provide a fuller picture of the long-term rewards for defensive excellence in the NBA.

